A Gentle Introduction and Application of Feature-Based Clustering with Psychological Time Series

### 2023-08-02 latex-results/07ClusterEvaluationInterpretation.tex
internal feedback lfb

Finally, we can also assess the clusters across other individual
difference variables \citep[e.g.,][]{monden2022}. This out-of-feature
comparison allows us to check for data artifacts, as well as check
whether the developmental clusters are associated with important social
markers and individual differences. To illustrate artifact checks, we
added the number of ESM measurements into the comparison and find that
the participants in the deterioration cluster (cluster 2) on average
completed slightly more ESM surveys in general and reported on more
intergroup interactions in particular (see \(n\) in
\fgrref[B]{fig:clusterFeatVar}). In our data exclusion procedures, we
ensured that the general time frame and completion rates are similar for
all participants and indeed the numbers in ESM measurements generally
are largely similar (e.g., see \(n\) for well-being and outgroup
attitudes). However, the difference in the reported number of
interactions might indicate either a clustering artifact or a meaningful
difference. The higher average number of interactions in cluster 2
could, for example, indicate a clustering artifact if variances are more
inflated due to the larger samples \citep{kogan2006}. In our case this
seems less likely because one out of four variables did not differ in
terms of the MAD (see \fgrref{fig:cluster_comparison_mad}). At the same
time, however, the difference in the number of experienced interactions
might also indicate a meaningful difference, where the deteriorating
cluster (cluster 2) on average reported more outgroup interactions
(\(difference\) = 1.028, \(t\)(150.83) = 7.50, \(p\) \textless{} .001,
\(95%CI [0.76, 1.30]), but these interactions were less voluntary (
\)difference\$ = -1.04, \(t\)(108.89) = -7.71, \(p\) \textless{} .001,
\(95%CI [-1.31, -0.77]), less meaningful (
\)difference\$ = -1.003, \(t\)(136.40) = -7.16, \(p\) \textless{} .001,
\(95%CI [-1.28, -0.73]), and less positive (
\)difference\$ = -1.38, \(t\)(152.31) = -11.94, \(p\) \textless{} .001,
\$95\%CI {[}-1.61, -1.15{]}). Such a finding is in line with past
research highlighting the role of negative intergroup interactions in
explaining intergroup relations
\citep[e.g.,][]{Barlow2012, Prati2021, Graf2014}.

@@

To further illustrate the utility of assessing out-of-feature individual
differences, we also compare the two samples in terms of the
participants' self-reported discrimination experiences in the
Netherlands (measured during the post-measurement). When looking at the
group comparison, we find that participants in the deteriorating cluster
(cluster 2) reported substantially higher levels of everyday
discrimination (\fgrref[B]{fig:clusterFeatVar}). Thus, both intensive
longitudinal (e.g., the sum of specific ESM measurements) and
cross-sectional variables (e.g., general discrimination differences)
that were not included in the original clustering step can be used to
explore and understand the cluster differences in more detail.

>>>>

Finally, we can also assess the clusters across other individual
difference variables \citep[e.g.,][]{monden2022}. This out-of-feature
comparison allows us to check for data artifacts, as well as check
whether the developmental clusters are associated with important social
markers and individual differences. To illustrate artifact checks, we
added the number of ESM measurements into the comparison and find that
the participants in the deterioration cluster (cluster 2) on average
completed slightly more ESM surveys in general and reported on more
intergroup interactions in particular (see \(n\) in
\fgrref[B]{fig:clusterFeatVar}). In our data exclusion procedures, we
ensured that the general time frame and completion rates are similar for
all participants and indeed the numbers in ESM measurements generally
are largely similar (e.g., see \(n\) for well-being and outgroup
attitudes). However, the difference in the reported number of
interactions might indicate either a clustering artifact or a meaningful
difference. The higher average number of interactions in cluster 2
could, for example, indicate a clustering artifact if variances are
substantially larger due to the larger samples
\citep[e.g., restriction of range in the smaller sample][]{kogan2006}.
In our case, this seems less likely because one out of four variables
did not differ in terms of the MAD (i.e., our selected measurement of
the time series variance; see \fgrref{fig:cluster_comparison_mad} for an
illustration). At the same time, however, the difference in the number
of experienced interactions might also indicate a meaningful difference,
where the deteriorating cluster (cluster 2) on average reported more
outgroup interactions (\textit{difference} = 1.03, \(t\)(150.83) = 7.50,
\(p\) \textless{} .001, \textit{95\%CI} {[}0.76, 1.30{]}), but these
interactions were less voluntary (\textit{difference} = -1.04,
\(t\)(108.89) = -7.71, \(p\) \textless{} .001, \textit{95\%CI} {[}-1.31,
-0.77{]}), less meaningful (\textit{difference} = -1.00, \(t\)(136.40) =
-7.16, \(p\) \textless{} .001, \textit{95\%CI} {[}-1.28, -0.73{]}), and
less positive (\textit{difference} = -1.38, \(t\)(152.31) = -11.94,
\(p\) \textless{} .001, \textit{95\%CI} {[}-1.61, -1.15{]}). Thus, while
more research is needed for a conclusive test, our data seems to suggest
that the differences in reported interactions are a meaningful
difference between the clusters. Such a finding would also be in line
with past research highlighting the role of negative intergroup
interactions in explaining intergroup relations
\citep[e.g.,][]{Barlow2012, Prati2021, Graf2014}.

@@

To further illustrate the utility of assessing out-of-feature individual
differences, we also compare the two samples in terms of the
participants' self-reported discrimination experiences in the
Netherlands (measured during the post-measurement). When looking at the
group comparison, we find that participants in the deteriorating cluster
(cluster 2) reported substantially higher levels of everyday
discrimination (\textit{difference} = 0.40, \(t\)(151.71) = 2.56, \(p\)
= 0.011, \textit{95\%CI} {[}0.09, 0.71{]};
\fgrref[B]{fig:clusterFeatVar}). Thus, both intensive longitudinal
(e.g., the sum of specific ESM measurements) and cross-sectional
variables (e.g., general discrimination differences) that were not
included in the original clustering step can be used to explore and
understand the cluster differences in more detail.

### 2023-08-02 sections/07ClusterEvaluationInterpretation.Rmd
internal feedback lfb

results$ar01$AttitudesDutch$conf.int

@@

results_df <- results_df %>%
      mutate(star = ifelse(p_value < 0.001, "***", ifelse(p_value < 0.01, "**", ifelse(p_value < 0.05, "*", ""))))

feature_plot_list <- list()

library(forestplot)
for (feature in unique(results_df$feature)) {
  # Subset the data for the current feature
  feature_data <- results_df[results_df$feature == feature, ]

  # Create a matrix for the forestplot function
  plot_matrix <- rbind(c("", "Difference", "Lower CI", "Upper CI"),
                       cbind(feature_data$concept,
                             feature_data$difference,
                             feature_data$lwr,
                             feature_data$upr))

  # Create the forest plot
  feature_plot_list[[feature]] <- forestplot::forestplot(labeltext = plot_matrix[, 1], 
             mean = as.numeric(plot_matrix[, 2]),
             lower = as.numeric(plot_matrix[, 3]),
             upper = as.numeric(plot_matrix[, 4]),
             xlab = "Difference",
             title = paste0("Forest plot for ", feature))
}

concept_plot_list <- list()
for (concept in unique(results_df$concept)) {
  # Subset the data for the current feature
  concept_data <- results_df[results_df$concept == concept, ]

  # Create a matrix for the forestplot function
  plot_matrix <- rbind(c("", "Difference", "Lower CI", "Upper CI"),
                       cbind(feature_data$concept,
                             feature_data$difference,
                             feature_data$lwr,
                             feature_data$upr))

  # Create the forest plot
  concept_plot_list[[concept]] <- forestplot::forestplot(labeltext = plot_matrix[, 1], 
             mean = as.numeric(plot_matrix[, 2]),
             lower = as.numeric(plot_matrix[, 3]),
             upper = as.numeric(plot_matrix[, 4]),
             xlab = "Difference")
}

dev.off()
pdf("figures/feature_comparison_mad.pdf")
feature_plot_list$mad
dev.off()
png("figures/feature_comparison_mad.png")
feature_plot_list$mad
dev.off()

@@

To further illustrate the utility of assessing out-of-feature individual differences, we also compare the two samples in terms of the participants' self-reported discrimination experiences in the Netherlands (measured during the post-measurement). When looking at the group comparison, we find that participants in the deteriorating cluster (cluster 2) reported substantially higher levels of everyday discrimination (\hl{T-TEST HERE!!}; \fgrref[B]{fig:clusterFeatVar}). Thus, both intensive longitudinal (e.g., the sum of specific ESM measurements) and cross-sectional variables (e.g., general discrimination differences) that were not included in the original clustering step can be used to explore and understand the cluster differences in more detail.

>>>>

@@

results_df <- results_df %>%
      mutate(
        star = ifelse(p_value < 0.001, "***", ifelse(p_value < 0.01, "**", ifelse(p_value < 0.05, "*", "")))
      ) %>% 
 left_join(
   variableLab,
   by = c('concept' = 'variable')
 )

feature_plot_list <- list()
for (feature in unique(results_df$feature)) {
  # Subset the data for the current feature
  feature_data <- results_df[results_df$feature == feature, ]

  df_plot <- data.frame(label = feature_data$label,
                        difference = feature_data$difference,
                        lwr = feature_data$lwr,
                        upr = feature_data$upr)

  feature_plot_list[[feature]] <- ggplot(df_plot, aes(y = label)) +
    geom_point(aes(x = difference), shape=15, size=2) +
    geom_linerange(aes(xmin = lwr, xmax = upr)) +
    geom_vline(xintercept = 0, linetype="dashed") +
    scale_x_continuous(expand = expansion(mult = 0)) +
    theme_Publication() + 
    theme(panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(), 
          axis.text.y = element_text(hjust=0), 
          axis.line.y = element_blank(),
          axis.ticks.y = element_blank(),
          axis.title.y = element_blank())
}

# Check if any graphics devices are open
if (length(dev.list()) > 0) {
  # Close the graphics device
  dev.off()
}
pdf("figures/feature_comparison_mad.pdf")
feature_plot_list$mad
dev.off()
png("figures/feature_comparison_mad.png")
feature_plot_list$mad
dev.off()

@@

To further illustrate the utility of assessing out-of-feature individual differences, we also compare the two samples in terms of the participants' self-reported discrimination experiences in the Netherlands (measured during the post-measurement). When looking at the group comparison, we find that participants in the deteriorating cluster (cluster 2) reported substantially higher levels of everyday discrimination (`r feat_ttest_apa(results_df, "exWB", "EvDayDiscr.post.z")`; \fgrref[B]{fig:clusterFeatVar}). Thus, both intensive longitudinal (e.g., the sum of specific ESM measurements) and cross-sectional variables (e.g., general discrimination differences) that were not included in the original clustering step can be used to explore and understand the cluster differences in more detail.
